pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-08-04 bmc_article.tex
changed supp fig links

Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Supplementary Figure 1 for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide practical recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Supplementary methods). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \cite{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 4). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \cite{tianMixology2018}, using SNP genotypes as the ground truth. For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \cite{kangMultiplexedDemuxlet2018}. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types). These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g., SNPs). We therefore do not expect a perfect accuracy in datasets involving cells of the same type across individuals (as in the demuxlet dataset).

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \cite{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties. This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells. Supplementary Figure 5 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 6), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 7). We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figure 8-10). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified{\color{red}, although such a pattern was not observed in the other datasets.}

We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{kuhnHungarianMethod1955} and monitored the mean F1 score across subpopulations. A first observation was that, although more stringent filtering tended to be associated with an increase in accuracy, it tended to plateau and could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12A). This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. In contrast, too lenient filtering (or lack thereof) can lead to misclassified cells, as was the case for the Zhengmix4 datasets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (\textit{default} - see methods) provided a good trade-off, often leading to high clustering performance (e.g. simMix and Zhengmix4 datasets) while retaining most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). PCA-based selection did not appear superior to feature-wise filters. We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B). Of note, doublet removal in conjunction with filtering tended to improve accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (e.g. FACS-sorted datasets). Indeed, the datasets where doublet removal did not have a clear positive impact (e.g. Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected. Overall, we therefore recommend the use of doublet removal followed by our \textit{default} filtering (see Methods) or a similar approach.

@@

Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \cite{freytagComparison2018}. We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 13). Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations. Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.

@@

An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}. However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences. We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:norm}A and Supplementary Figure 14). Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{sctransform}. Surprisingly, however, regressing out the covariates tended to increase the association with it.

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that some datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods. {\color{red}An exception was \textit{scVI}-based normalization, which performed similarly to other methods on some datasets (Kumar, simMix2, Zhengmix4eq, Zhengmix4uneq), but very poorly on others, in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; of note, these values were not meant to be used for clustering, and the usage of the \textit{scVI} latent space will be discussed below along with dimensionality reductions)}. Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal or poorer performance on MI, ARI and ARI at true number of clusters, with a strongest negative impact among the Smart-seq datasets (Koh and Kumar). Regressing out covariates using \textit{sctransform} instead, tended to have a much milder effect, possibly owing to its regularization. In general, \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters. Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 16). We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 17), at little apparent cost on computation time (Supplementary Figure 18).

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 19A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figure 20-21). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figure 20-22). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figure 20-21). \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:selection}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets (Figure \ref{fig:selection}B). Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also tended to lead to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).

@@

The impact of the dimensionality reduction methods tested was far greater than that of normalization or feature selection (Supplementary Figure 23). Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. {\color{red} Like GLM-PCA, \textit{scVI}'s Linearly Decoded (LD) data \cite{SvenssonLDVAE2020} and latent space do not explicitly rely on normalized counts. Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Supplementary Figure 15). In terms of clustering accuracy, \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).}

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figure 16 and 24). This tended to affect the accuracy of the clustering (Supplementary Figure 25), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

{\color{red} \textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Supplementary Figure 26). Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Supplementary Figure 27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D). \textit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violates the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others. We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by \textit{MCVR} and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.

@@

Since ARI is dominated by differences in the number of clusters (Supplementary Figure 2-3) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}). MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 3). Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters. We also tracked the mean F1 score and the ARI at the true number of clusters.

>>>>

Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Additional File 1: Fig S1 for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide practical recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Supplementary methods). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Additional File 1: Fig S2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \cite{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Additional File 1: Fig S4). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \cite{tianMixology2018}, using SNP genotypes as the ground truth. For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \cite{kangMultiplexedDemuxlet2018}. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types). These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g., SNPs). We therefore do not expect a perfect accuracy in datasets involving cells of the same type across individuals (as in the demuxlet dataset).

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \cite{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties. This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells. Additional File 1: Fig S5 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Additional File 1: Fig S6), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Additional File 1: Fig S7). We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Additional File 1: Fig S8-10). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified{\color{red}, although such a pattern was not observed in the other datasets.}

We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Additional File 1: Fig S11). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{kuhnHungarianMethod1955} and monitored the mean F1 score across subpopulations. A first observation was that, although more stringent filtering tended to be associated with an increase in accuracy, it tended to plateau and could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Additional File 1: Fig S12A). This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. In contrast, too lenient filtering (or lack thereof) can lead to misclassified cells, as was the case for the Zhengmix4 datasets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (\textit{default} - see methods) provided a good trade-off, often leading to high clustering performance (e.g. simMix and Zhengmix4 datasets) while retaining most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). PCA-based selection did not appear superior to feature-wise filters. We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Additional File 1: Fig S12B). Of note, doublet removal in conjunction with filtering tended to improve accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (e.g. FACS-sorted datasets). Indeed, the datasets where doublet removal did not have a clear positive impact (e.g. Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected. Overall, we therefore recommend the use of doublet removal followed by our \textit{default} filtering (see Methods) or a similar approach.

@@

Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \cite{freytagComparison2018}. We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Additional File 1: Fig S13). Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations. Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.

@@

An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}. However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences. We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:norm}A and Additional File 1: Fig S14). Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{sctransform}. Surprisingly, however, regressing out the covariates tended to increase the association with it.

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that some datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods. {\color{red}An exception was \textit{scVI}-based normalization, which performed similarly to other methods on some datasets (Kumar, simMix2, Zhengmix4eq, Zhengmix4uneq), but very poorly on others, in terms of both the subpopulation silhouette and clustering accuracy (Additional File 1: Fig S13; of note, these values were not meant to be used for clustering, and the usage of the \textit{scVI} latent space will be discussed below along with dimensionality reductions)}. Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal or poorer performance on MI, ARI and ARI at true number of clusters, with a strongest negative impact among the Smart-seq datasets (Koh and Kumar). Regressing out covariates using \textit{sctransform} instead, tended to have a much milder effect, possibly owing to its regularization. In general, \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters. Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Additional File 1: Fig S16). We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Additional File 1: Fig S17), at little apparent cost on computation time (Additional File 1: Fig S18).

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Pearson correlation >0.9, Additional File 1: Fig S19A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Additional File 1: Fig S19B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Additional File 1: Fig S20-21). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Additional File 1: Fig S20-22). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Additional File 1: Fig S20-21). \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:selection}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets (Figure \ref{fig:selection}B). Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also tended to lead to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Additional File 1: Fig S22).

@@

The impact of the dimensionality reduction methods tested was far greater than that of normalization or feature selection (Additional File 1: Fig S23). Overall, weighting the components by their explained variance tended to increase silhouette width (Additional File 1: Fig S23A) and, in most datasets, clustering accuracy (Additional File 1: Fig S23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. {\color{red} Like GLM-PCA, \textit{scVI}'s Linearly Decoded (LD) data \cite{SvenssonLDVAE2020} and latent space do not explicitly rely on normalized counts. Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Additional File 1: Fig S15). In terms of clustering accuracy, \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).}

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Additional File 1: Fig S16 and 24). This tended to affect the accuracy of the clustering (Additional File 1: Fig S25), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

{\color{red} \textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Additional File 1: Fig S26). Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Additional File 1: Fig S27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D). \textit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violates the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others. We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by \textit{MCVR} and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.

@@

Since ARI is dominated by differences in the number of clusters (Additional File 1: Fig S2-3) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}). MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Additional File 1: Fig S3). Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters. We also tracked the mean F1 score and the ARI at the true number of clusters.
